Model for DES Type Ia Supernova Cosmology Analysis

### 2015-08-15 doc/model.tex
making results of 50% run

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   The forward modeling
   flow goes from left to right, starting with the description of the universe, observatory,
   and data.    A transient has host galaxy $g_i$, which determines its redshift $z_i$
   and galaxy parameters $\theta_{gi}$.
   A model with parameters $\theta_\mu$ fix the distance modulus $\mu_i$.
   The transient type $T$ depends on the host-galaxy parameters  $\theta_{gi}$
   and rates $\theta_r$.   The transient
   parameters $\theta_T^X$, $\theta_{Ti}^X$ (and perhaps the host galaxy) determine the luminosity $L$.       The 
   incoming photon flux $n$, $n_g$  are then fixed
   with redshift and distance modulus.
   The instrumental transmission function $\phi$ is calibrated with data ${Z}$ and
   gives the expected
   counts $\overline{\mathit{ADU}}$, $\overline{\mathit{ADU}}_g$. 
   The transient has realized light curves (${ADU}$) and passes selection criteria
   through $\tau_i$ to be included in the sample.  Some transients have spectral data
   ${X}_S$.  The properties of the galaxy identified as the host galaxy is $\theta_G$. 
   \label{pgm:fig}}
\end{figure}

@@

The transients are divided into samples defined by available data: {\it obs} are transients observed spectroscopically while active, from which $T_S=1$ are those typed
as SN~Ia and $T_S=0$ as non-Ia.  {\it mis} transients are not observed
spectroscopically when active.

\begin{itemize}
\item All transients have measured counts at peak brightness $\mathit{ADU}$.
For transients in set {\it mis} without active spectroscopy:
\begin{multline}
p(\mathit{ADU}| z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})=\\
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right)+\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
The luminosity distance $d_L$ is for a flat universe.  
\item The transients of subset {\it obs} have spectroscopic observations of type $T_S$ and redshift
$z_S$.
\begin{itemize}
\item A direct association of $z_S=z$ is made.
\begin{equation}
p(z_S|z) = \delta(z_S-z)
\label{specz:eqn}
\end{equation}
\item For transients in set $T_S=1$:
\begin{equation}
p(\mathit{ADU}, T_S=1 | z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia})=
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right).
\end{equation}
\item For transients in set $T_S=0$:
\begin{multline}
p(\mathit{ADU}, T_S=0 | z, \Omega_M, w, \theta_r, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})= \\
\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
\end{itemize}
\item All transients are associated with a host galaxy with spectroscopic redshift $\theta_G$,
\begin{equation}
p(\theta_G|z) =
	P_{host} \ln{\mathcal{N}}(\theta_G;\ln(1+z),0.001) + (1-P_{host})\left(\frac{3 \theta_G^2}{z_{max}^3 - z_{min}^3}\right).
\end{equation}
The host is determined correctly with probability $P_{host}=0.98$ in which case the
host redshift is determined from a log-normal distribution centered around the true redshift
with 0.001 fractional uncertainty.   Otherwise the redshift is drawn from the ``uniform in volume'' term on the right.
[Note: to speed the cacluation, I only consider host-galaxy measurements
of the {\it mis} set.]
\item This simple example does not consider data $\theta_S$, $\tau$ or $Z$, i.e.
spectroscopic parameters, sample-selection effects, and calibration.
\end{itemize}

@@

\subsection{Analysis of Simulated Data}
Simulated data is drawn from the above model with the following parameter
choices:
\begin{itemize}
\item Cosmological parameters $\Omega_M=0.28$, $w=-1$.
\item Type~Ia supernovae with intrinsic dispersion $\sigma_{Ia}=0.1$.
\item Non-Ia supernovae 2 magnitudes fainter than SNe~Ia $\alpha_{\mathit{non-Ia}} = 10^{-2/2.5} \alpha_{Ia}$ with 
 $\sigma_{\mathit{non-Ia}}=1$.
\item $\theta_r=0.8$.
\item 200 supernovae uniformly distributed in the range $0.1<z<1$.
\item $z_{min}=0.05$ and $z_{max}=1.5$ used for the range of possible true redshifts.
\end{itemize}
From a realization of these data, spectroscopic follow-up is
accumulated: analysis is performed at 50\%, 75\%, and 100\% 
completeness.

@@

%\subsection{Type~Ia Supernova Model}
%\label{snmodel:sec}
%The DES light curve analysis in principle may be used for typing and determining
%redshift, in addition to getting supernova properties.   
%Generally the relevant term is
%\begin{equation}
%P({D}_i=1,{\mathit{ADU}}_{i}, {T}_{Si}, {z}_{Si}, {\theta}_{Si}, {z}_{Hi}, {\theta}_{Hi}
% | \mu_i,z_i,T_i,\theta_T,\theta_{Ti},\phi(\lambda)).
% \end{equation}
%Detection and typing depend on the realized light curves, and hence the truncated
%distribution of discovered transients typed as SN~Ia.
%The model contains a piece $P({\mathit{ADU}}_{i} {z}_{Si}| \mu_i,z_i,\theta_T, \theta_{Ti}, ,\phi(\lambda))$, which
%is the piece that describes underlying SN~Ia properties and light-curve fits.
%\subsection{Detection Efficiency}
%There term $P({D}|\mathit{ADU}_V)$.  For all high-redshift supernova searches
%I know of, this depends on the signal-to-noise of the data, and very weakly on the
%structure of the underlying galaxy.  Our detection is now based on machine learning,
%we should confirm that detection dependencies do not depend on our model parameters
%(except for RA/Dec).

@@

%\section{Misc.}
%DES plans a ``photometry''-only analysis, in that a significant subset of transients used
%in the Hubble diagram will not observed spectroscopically while active.  Therefore,
%our model must account for transient mistyping and host-galaxy mismatches.  To obtain
%our targeted
%accuracy we need to account for covariances that occur due to photometric
%calibration and Malmquist bias.  Our model will contain elements that may not
%have been included in previous analyses. 

@@

%Here $\mu$ is modeled; most analyses instead
%directly have cosmological parameters $\{\Omega_M, \Omega_\Lambda, w_0, w_a\}$.
%A motivation is to provide results that can be applied to test arbitrary cosmological models:
%indeed the model-dependence of the $\mu$'s reported for high-$z$ supernovae 
%\citep[e.g.][]{2012ApJ...746...85S} is a source of confusion to the community.
%%An example of a non-parametric model for $\mu$ that is independent of cosmological
%%parameters is given in the Appendix of
%%\citet{2013ApJ...764..116W}.

>>>>

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.pdf} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   The forward modeling
   flow goes from left to right, starting with the description of the universe, observatory,
   and data.    A transient has host galaxy $g_i$, which determines its redshift $z_i$
   and galaxy parameters $\theta_{gi}$.
   A model with parameters $\theta_\mu$ fix the distance modulus $\mu_i$.
   The transient type $T$ depends on the host-galaxy parameters  $\theta_{gi}$
   and rates $\theta_r$.   The transient
   parameters $\theta_T^X$, $\theta_{Ti}^X$ (and perhaps the host galaxy) determine the luminosity $L$.       The 
   incoming photon flux $n$, $n_g$  are then fixed
   with redshift and distance modulus.
   The instrumental transmission function $\phi$ is calibrated with data ${Z}$ and
   gives the expected
   counts $\overline{\mathit{ADU}}$, $\overline{\mathit{ADU}}_g$. 
   The transient has realized light curves (${ADU}$) and passes selection criteria
   through $\tau_i$ to be included in the sample.  Some transients have spectral data
   ${X}_S$.  The properties of the galaxy identified as the host galaxy is $\theta_G$. 
   \label{pgm:fig}}
\end{figure}

@@

The transients are divided into samples defined by available data: {\it obs} are transients observed spectroscopically while active, from which $T_S=1$ are those typed
as SN~Ia and $T_S=0$ as non-Ia.  {\it mis} transients are not observed
spectroscopically when active.  The same SN~Ia and non-Ia parameters are taken to
apply to both {\it obs} and {\it mis} samples.

\begin{itemize}
\item All transients have measured counts at peak brightness $\mathit{ADU}$.
For transients in set {\it mis} without active spectroscopy:
\begin{multline}
p(\mathit{ADU}| z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia}, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})=\\
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right)+\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
The luminosity distance $d_L$ is for a flat universe.  
\item The transients of subset {\it obs} have spectroscopic observations of type $T_S$ and redshift
$z_S$.
\begin{itemize}
\item A direct association of $z_S=z$ is made.
\begin{equation}
p(z_S|z) = \delta(z_S-z)
\label{specz:eqn}
\end{equation}
\item For transients in set $T_S=1$:
\begin{equation}
p(\mathit{ADU}, T_S=1 | z, \Omega_M, w, \theta_r, \alpha_{Ia},\sigma_{Ia})=
\theta_r\mathcal{N}\left(\frac{\alpha_{Ia}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{Ia}\right).
\end{equation}
\item For transients in set $T_S=0$:
\begin{multline}
p(\mathit{ADU}, T_S=0 | z, \Omega_M, w, \theta_r, \alpha_{\mathit{non-Ia}},\sigma_{\mathit{non-Ia}})= \\
\left(1-\theta_r\right)
\mathcal{N}\left(\frac{\alpha_{\mathit{non-Ia}}}{d_L^2(z;\Omega_M, w)}, \frac{\ln{10}}{2.5}\sigma_{\mathit{non-Ia}}\right).
\end{multline}
\end{itemize}
\item All transients are associated with a host galaxy with spectroscopic redshift $\theta_G$,
\begin{equation}
p(\theta_G|z) =
	P_{host} \ln{\mathcal{N}}(\theta_G;\ln(1+z),0.001) + (1-P_{host})\left(\frac{3 \theta_G^2}{z_{max}^3 - z_{min}^3}\right).
\end{equation}
The host is determined correctly with probability $P_{host}$ in which case the
host redshift is determined from a log-normal distribution centered around the true redshift
with 0.001 fractional uncertainty.   Otherwise the redshift is drawn from the ``uniform in volume'' term on the right.
[Note: to speed the cacluation, I only consider host-galaxy measurements
of the {\it mis} set.]
\item This simple example does not consider data $\theta_S$, $\tau$ or $Z$, i.e.
spectroscopic parameters, sample-selection effects, and calibration.
\end{itemize}

@@

\subsection{Analysis of Simulated Data}
Simulated data is drawn from the above model with the following parameter
choices:
\begin{itemize}
\item Cosmological parameters $\Omega_M=0.28$, $w=-1$.
\item Type~Ia supernovae with intrinsic dispersion $\sigma_{Ia}=0.1$.
\item Non-Ia supernovae 2 magnitudes fainter than SNe~Ia $\alpha_{\mathit{non-Ia}} = 10^{-2/2.5} \alpha_{Ia}$ with 
 $\sigma_{\mathit{non-Ia}}=1$.
\item $\theta_r=0.8$.
\item 200 supernovae uniformly distributed in the range $0.1<z<1$.
\item $z_{min}=0.05$ and $z_{max}=1.5$ used for the range of possible true redshifts.
\item $P_{host}=0.98$.
\end{itemize}
From a realization of these data, spectroscopic follow-up is
accumulated: analysis is performed at 50\%, 75\%, and 100\% 
completeness.

The simulated data is analyzed using \citet{pystan-software:2015},
the Python interface to
\citet{stan-software:2015}, a probabilistic programming language for
inference of Bayesian models with Hamiltonian Monte Carlo.  There are
four chains run, each with 50,000 samples then thinned by 50.

\subsubsection{50\% Spectroscopic Confirmation}
Results for the $\Omega_M$--$w$ posterior for the
case where 50\% of the transients have a direct spectrum
are shown in Figure~\ref{contour.100:fig}.
As will be discussed shortly, the Stan run does not result in a Markov Chain Monte Carlo
that has
converged to a stationary solution.  Nevertheless, the results are presented here
in the hope that they will spark progress in producing robust posteriors or at the
least provide an alert to the computational challenge of the analysis.
Despite their dubious quality, we can be consoled that the distributions
look similar to those of JLA \citep{2014A&A...568A..22B} 
and so satisfy qualitative expectations.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.6]{/Users/akim/project/abc/results/{contour.100}.pdf} 
   \caption{Probability distribution function for $\Omega_M$--$w$  when 50\% of transients
   get spectra while active.
   Solid lines represent the input cosmology.  Dashed lines in the
   histograms are the 0.16 and 0.86 quantiles. 
   \label{contour.100:fig}}
\end{figure}

A transient
without a spectrum has uncertain type and uncertain redshift: the redshift
of the associated host galaxy, and redshifts associated with SNe~Ia and non-Ia's
at the observed magnitude, are viable but disjoint possibilities.  The initial condition
for redshifts of transients without an active spectrum is near the true redshift
for chains 0 and 2, and near the host redshift in chains 1 and 3.

Figure~\ref{missed_wrong.100:fig} shows draws
for the redshift parameter for two transients: one is Type~Ia and the other is
non-Ia, neither have a spectrum and both have an incorrect host galaxy association.
The SN~Ia is placed at $z=0.95$ but is misassociated with a galaxy at
$z=1.24$.  The posterior shows a preference for the input redshift, but as the analyzer
must entertain the possibility it is non-Ia the posterior sometimes
goes to lower redshifts. Although two of the chains start near the incorrect-host redshift,
there are no such redshifts drawn.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.6]{/Users/akim/project/abc/results/{missed_wrong.100}.pdf} 
   \caption{Redshift posterior for a Type~Ia and a non-Ia that were not observed
   spectroscopically and for whom the host-galaxy is misassociated.
   The results of the four chains are
   placed in sequence.   \label{missed_wrong.100:fig}}
\end{figure}

The non-Ia is placed at $z=0.39$ and is misassociated with a galaxy at $z=0.48$.
Three of the four chains stay in the neighborhood of the incorrect-host redshift.
Chain 2, whose initial condition was at the true redshift, also accepts redshift solutions as
if were a Type~Ia.  Clearly this chain does not mix with the others.

Figure~\ref{posterior.100:fig} shows the log-posteriors and histograms of
the $w$-draws from Chain 2 compared to the others.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[scale=0.5]{/Users/akim/project/abc/results/{posterior.100}.pdf} 
   \includegraphics[scale=0.5]{/Users/akim/project/abc/results/{wsub.100}.pdf} 
   \caption{Log-posterior and $w$-draws from  Chain 2 compared to the others.  \label{posterior.100:fig}}
\end{figure}

@@

@@

@@
